---
title: "Evaluating Family Engagement and Disparities in Health Information Access through Patient Portal Use within a NICU"
subtitle: "BMIN503 Final Project"
author: "Osvaldo Mercado"
date: "`r Sys.Date()`"
format: 
  html:
    fig-width: 8
    fig-height: 6
    toc: true
    toc-location: left
    css: "https://github.research.chop.edu/pages/CQI/chop-bootstrap/bootstrap-3/bootstrap.min.css"
editor: visual
embed-resources: true
editor_options: 
  chunk_output_type: inline
execute: 
  warning: false
  message: false
---

```{=html}
<style>
#TOC {
  background-color: #f9f9f9; /* Light gray background */
  border: 1px solid #ddd;   /* Add a subtle border */
  padding: 10px;            /* Add padding inside the box */
  border-radius: 5px;       /* Rounded corners */
  box-shadow: 2px 2px 5px rgba(0, 0, 0, 0.1); /* Optional shadow for depth */
}
</style>
```

# Overview

This project aims to analyze how families in the neonatal intensive care unit (NICU) engage with their infant's patient portal or MyCHOP accounts (an online electronic health record) to access various aspects of their child’s care. The project will examine patterns of patient portal use, including frequency, types of information accessed, and timing, with a focus on understanding how these interactions vary across demographics and socioeconomic factors.

# Introduction

Families with infants in the NICU experience high levels of stress and complexity in navigating their child’s healthcare. Access to MyCHOP, an online portal for EHR access, provides a valuable resource for families to view updates on their child’s treatment, review provider notes, and stay informed about care decisions. However, disparities may exist in how effectively families from different backgrounds can use this tool to stay informed and engaged. Understanding how families access their MyCHOP accounts, what types of data they prioritize, and when they engage with the portal could reveal insights into potential barriers to access or disparities in engagement. These insights are especially crucial in the NICU setting, where clear communication and access to information are critical for family-centered care. This study will analyze usage patterns among families, considering factors such as race, ethnicity, median household income, and property values, to identify trends and inform interventions that could enhance equitable access to critical health information.

# Methods

The dataset for this project was sourced from the Children's Hospital of Philadelphia (CHOP) Clinical Data Warehouse. It was retrieved from tables stored within Clarity and the Trusted Data Layer. As this dataset is not publicly accessible, the data tables cannot be shared. This project serves as an exploratory analysis of the variables described earlier. The methods section begins with loading the required packages for this analysis.

## Install packages

```{r}
#| eval: false

install.packages("readr")
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("lubridate")
install.packages("gtsummary")
install.packages("RColorBrewer")
install.packages("cowplot")
install.packages("dplyr")

```

## Load packages

```{r}
library(readr) # Data import
library(tidyverse) # Data manipulation, visualizaion and analysis
library(ggplot2) # For creation of graphs
library(lubridate) # For manipulating dates
library(gtsummary) # Summary statistics
library(RColorBrewer) # For personalization of color
library(cowplot) # For combining multiple plots into one
library(dplyr) # For data cleaning and filtering
library(forcats)  # Allows re-coding of factors

```

## Upload data

```{r}
#Upload patient data
patient_data <- read.csv(file.choose())
message_data <- read.csv(file.choose())

```

## Exploration and Manipulation of patient data table

```{r}
# Count total unique IDs, gives unique patients
total_unique <- length(unique(patient_data$MRN))

# Count occurrences of repeated IDs, gives sum of repeat patient encounters
repeated <- sum(table(patient_data$MRN) > 1)

# Print results
cat("Total unique IDs:", total_unique, "\nNumber of repeated IDs:", repeated, "\n")

```

```{r}
# Sort the data by MRN and admission_date
patient_data_ordered <- patient_data[order(patient_data$MRN, patient_data$HOSPITAL_ADMIT_DATE), ]

# Keep only the first occurrence of each MRN
initial_admissions <- patient_data_ordered[!duplicated(patient_data_ordered$MRN), ]

#Keep only data for admissions prior to September 30, 2024 (available messaging data)
initial_admissions <- initial_admissions[initial_admissions$HOSPITAL_ADMIT_DATE < as.Date("2024-09-30"), ] 

# Count total records in the filtered table
total_records_initial <- nrow(initial_admissions)

# Display the total count
cat("Total records in the intial admission table:", total_records_initial, "\n") #Equivalent count to unique ID's above
```

```{r}
# Calculate total gestational age in TOTAL days: combination of two columns
initial_admissions$GESTATIONAL_AGE_TOTAL_DAYS <- 
  initial_admissions$GESTATIONAL_AGE_COMPLETE_WEEKS * 7 + #multiply by 7 days in week
  initial_admissions$GESTATIONAL_AGE_REMAINDER_DAYS

# Calculation of average GA in TOTAL days
average_GA_days <- mean(na.omit(initial_admissions$GESTATIONAL_AGE_TOTAL_DAYS))

# Convert the average back to weeks and days for interpretation
average_weeks <- floor(average_GA_days / 7) #Using floor to round down weeks
average_days <- round(average_GA_days %% 7, 1)  #Using %% to obtain remainder when dividing by 7

# Combine the results
cat("Average Gestational Age:", average_weeks, "weeks and", average_days, "days\n")
```

To analyze the stored data effectively, the data table needs to be properly converted and formatted. Below are steps to prepare the data so it can be summarized and manipulated correctly using factors.

Sex Column Exploration and Conversion

```{r}
# Additional analysis and conversion of data stored: SEX
unique(initial_admissions$SEX) # Stored as M, F, U
initial_admissions <- initial_admissions |>  # Remove limited 3 U
  dplyr::filter(SEX != "U")
# Conversion to factor
initial_admissions$SEX <- as.factor(initial_admissions$SEX)
initial_admissions$SEX <- droplevels(initial_admissions$SEX)
table(initial_admissions$SEX)
```

Race Column Exploration and Conversion

```{r}
# Additional analysis and conversion of data stored: RACE
## Will combine/group certain variables for clarity and simplicity 
unique(initial_admissions$RACE) 
initial_admissions <- initial_admissions |>
  dplyr::mutate(RACE = case_when(
      RACE %in% c("Refused", "Unknown", "Asked but unknown", "", "Choose not to disclose") ~ "Unknown/Missing/Refused",
      TRUE ~ RACE
    )
  )
# Conversion to factor
initial_admissions$RACE <- as.factor(initial_admissions$RACE)
table(initial_admissions$RACE)
```

Ethnicity Column Exploration and Conversion

```{r}
# Additional analysis and conversion of data stored: ETHNICITY
## Will combine/group certain variables for clarity and simplicity 
unique(initial_admissions$ETHNICITY) 
initial_admissions <- initial_admissions |> 
  dplyr::mutate(ETHNICITY = case_when(
      ETHNICITY %in% c("Refused", "Unknown", "Asked But Unknown", "", "Choose Not To Disclose", "Other") ~ "Unknown/Missing/Refused",
      TRUE ~ ETHNICITY
    )
  )
# Conversion to factor
initial_admissions$ETHNICITY <- as.factor(initial_admissions$ETHNICITY)
table(initial_admissions$ETHNICITY)
```

Preferred Language Column Exploration and Conversion

```{r}
# Additional analysis and conversion of data stored: PREFERRED LANGUAGE
## Will combine/group certain variables for clarity and simplicity 
unique(initial_admissions$PREFERRED_LANGUAGE) 
# Group languages with fewer than 15 occurrences as "Other" and keep "Unknown" and NULL seperate
initial_admissions <- initial_admissions |>
  dplyr::mutate(PREFERRED_LANGUAGE = case_when(
      is.na(PREFERRED_LANGUAGE) | PREFERRED_LANGUAGE == "" ~ "Unknown/Missing",  # Combine NULL/missing and "" into Unknown/Missing
      PREFERRED_LANGUAGE == "UNKNOWN" ~ "Unknown/Missing",                       # Keep UNKNOWN in this group
      PREFERRED_LANGUAGE %in% names(which(table(PREFERRED_LANGUAGE) < 15)) ~ "Other",  # Group languages with <15 occurrences
      PREFERRED_LANGUAGE == "Other" ~ "Other",                          # Explicitly include "Other" in this group
      TRUE ~ PREFERRED_LANGUAGE                                 
    )
  )

# Conversion to factor
initial_admissions$PREFERRED_LANGUAGE <- as.factor(initial_admissions$PREFERRED_LANGUAGE)
table(initial_admissions$PREFERRED_LANGUAGE)

```

Payor Group Column Exploration and Conversion

```{r}

# Additional analysis and conversion of data stored: PAYOR GROUP
unique(initial_admissions$PAYOR_GROUP) 
#Recode Medical Assistance as Medicaid
initial_admissions$PAYOR_GROUP <- 
  fct_recode(initial_admissions$PAYOR_GROUP, "MEDICAID" = "MEDICAL ASSISTANCE")
initial_admissions <- initial_admissions |>
  dplyr::mutate(PAYOR_GROUP = if_else(is.na(PAYOR_GROUP) | PAYOR_GROUP == "", "Missing/Unknown", PAYOR_GROUP))

# Conversion to factor
initial_admissions$PAYOR_GROUP <- as.factor(initial_admissions$PAYOR_GROUP)
table(initial_admissions$PAYOR_GROUP)

```

Additional data manipulation (include dates, MyCHOP activation, MyCHOP declined, inborn)

```{r}
#Converting various column dates from "character" to "date"
initial_admissions <- initial_admissions |>
  mutate(
    DOB = as.Date(initial_admissions$DOB),
    HOSPITAL_ADMIT_DATE = as.Date(HOSPITAL_ADMIT_DATE),
    HOSPITAL_DISCHARGE_DATE = as.Date(HOSPITAL_DISCHARGE_DATE),
    EPISODE_START_DATE = as.Date(EPISODE_START_DATE),
    EPISODE_END_DATE = as.Date(EPISODE_END_DATE)
  )

#Exploration of MyCHOP Columns
# Count rows where MYCHOP_ACTIVATION_IND is 1
activation_count <- sum(initial_admissions$MYCHOP_ACTIVATION_IND == 1, na.rm = TRUE)

# Count rows where MYCHOP_DECLINED_IND is 1
declined_count <- sum(initial_admissions$MYCHOP_DECLINED_IND == 1, na.rm = TRUE)

# Display the results
cat("Number of activations (MYCHOP_ACTIVATION_IND = 1):", activation_count, 
    "\nNumber of declines (MYCHOP_DECLINED_IND = 1):", declined_count, "\n")

# Count the number of inborn patients (inborn_IND = 1)
inborn_count <- sum(initial_admissions$INBORN_IND == 1, na.rm = TRUE)
cat("Number of inborn patients:", inborn_count, "out of", total_unique, "\n")
```

## Exploration and Manipulation of Messaging Table

```{r}
#When I initially queried the messaging information table for my included patients, it contained 37 million rows of information. After some data cleaning and in the interest of simpliying my data utilization in RStudio, I filtered specifically for pta.ua_time BETWEEN fne.episode_start_date AND fne.episode_end_date to make sure that the results only occured while a patient was admitted during my time window.

# Count total unique rows, gives total number of messages provided in filtered data set
total_unique_messages <- nrow(message_data)

# Count total unique patients, gives total number of patients who had their MyChart accessed during NICU admission
total_unique_messages_pat <- length(unique(message_data$MRN))

# Display results
cat("Total unique messages:", total_unique_messages, "\nNumber of unique patients:", total_unique_messages_pat, "\n")

```

Filtering table to only include patient access to MyChart (remove ShareEveryhere and External_Client)

```{r}
# Mutating dates
message_data <- message_data |>
  mutate(UA_TIME = as.Date(message_data$UA_TIME))

# Table to show counts of different individuals accessing MyChart
table(message_data$UA_PERSON_TYPE_C)

#Filtering for only patient and user (proxy)
filtered_message_data <- message_data |>
  dplyr::filter(message_data$UA_PERSON_TYPE_C %in% c(1, 2))
  
```

Creation of a lookup table to understand the numeric values of MYC_UA_TYPE_C. This column uses a numeric value which I can then associate with the lookup table I created from the Epic User Web to understand the action being completed.

```{r}
# Upload data table including Epic released information about MyChart actions
epic_lookup_table <- read.csv(file.choose())
epic_lookup_table <- as_tibble(epic_lookup_table)
```

# Results

The initial part of this section contains descriptive statistics for patients admitted to the NICU from January 2018 to \*\*\*. Displayed below is a summary table for the overall patient data set.

```{r}
summary_table <- initial_admissions |>
  tbl_summary(
    include = c(SEX, RACE, ETHNICITY, PREFERRED_LANGUAGE, PAYOR_GROUP, 
                MYCHOP_ACTIVATION_IND, MYCHOP_DECLINED_IND, 
                GESTATIONAL_AGE_TOTAL_DAYS, NICU_LOS_DAYS, INBORN_IND),
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",  # Median (IQR) for continuous variables
      all_categorical() ~ "{n} ({p}%)"              # Count and percentage for categorical variables
    ),
    label = list(
      SEX ~ "Gender",
      RACE ~ "Race",
      ETHNICITY ~ "Ethnicity",
      PREFERRED_LANGUAGE ~ "Preferred Language",
      PAYOR_GROUP ~ "Insurance",
      MYCHOP_ACTIVATION_IND ~ "MyCHOP Activated",
      MYCHOP_DECLINED_IND ~ "MyCHOP Declined",
      GESTATIONAL_AGE_TOTAL_DAYS ~ "Gestational Age (Days)",
      NICU_LOS_DAYS ~ "NICU Length of Stay (Days)",
      INBORN_IND ~ "Inborn"
    )
  )

summary_table <- summary_table %>%
  modify_header(label = "**Variable**") %>%
  modify_header(stat_0 = "**N (%)**") %>%
  modify_caption("**Summary Table**") %>%
  bold_labels()

summary_table
```

# Conclusion

This the conclusion. The @sec-results can be invoked here.

## References
