---
title: "Evaluating Family Engagement and Disparities in Health Information Access through Patient Portal Use within a NICU"
subtitle: "BMIN503 Final Project"
author: "Osvaldo Mercado"
date: "`r Sys.Date()`"
format: 
  html:
    fig-width: 8
    fig-height: 6
    toc: true
    toc-location: left
    css: "https://github.research.chop.edu/pages/CQI/chop-bootstrap/bootstrap-3/bootstrap.min.css"
editor: visual
embed-resources: true
editor_options: 
  chunk_output_type: inline
execute: 
  warning: false
  message: false
---

# Overview

This project aims to analyze how families in the neonatal intensive care unit (NICU) engage with their infant's patient portal or MyCHOP accounts (an online electronic health record) to access various aspects of their child’s care. The project will examine patterns of patient portal use, including frequency, types of information accessed, and timing, with a focus on understanding how these interactions vary across demographics and socioeconomic factors.

# Introduction

Families with infants in the NICU experience high levels of stress and complexity in navigating their child’s healthcare. Access to MyCHOP, an online portal for EHR access, provides a valuable resource for families to view updates on their child’s treatment, review provider notes, and stay informed about care decisions. However, disparities may exist in how effectively families from different backgrounds can use this tool to stay informed and engaged. Understanding how families access their MyCHOP accounts, what types of data they prioritize, and when they engage with the portal could reveal insights into potential barriers to access or disparities in engagement. These insights are especially crucial in the NICU setting, where clear communication and access to information are critical for family-centered care. This study will analyze usage patterns among families, considering factors such as race, ethnicity, median household income, and property values, to identify trends and inform interventions that could enhance equitable access to critical health information.

# Methods

The dataset for this project was sourced from the Children's Hospital of Philadelphia (CHOP) Clinical Data Warehouse. It was retrieved from tables stored within Clarity and the Trusted Data Layer. As this dataset is not publicly accessible, the data tables cannot be shared. This project serves as an exploratory analysis of the variables described earlier. The methods section begins with loading the required packages for this analysis.

## Install packages

```{r}
#| eval: false

install.packages("readr")
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("lubridate")
install.packages("gtsummary")
install.packages("RColorBrewer")
install.packages("cowplot")
install.packages("dplyr")

```

## Load packages

```{r}
#| eval: false

library(readr)
library(tidyverse) 
library(ggplot2) # For graphs
library(lubridate) # For manipulating dates
library(gtsummary) #Summary statistics
library(RColorBrewer) # For coloring the plots
library(cowplot) # For combining multiple plots into one
library(dplyr) # For data cleaning

```

## Upload data

```{r}


```

# Results {#sec-results}

Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

# Conclusion

This the conclusion. The @sec-results can be invoked here.

# References
